---
title: "National Student Clearinghouse, 2012-2020"
subtitle: "State of College Report"
author: "Shannon Coulter"
date: "`r format(Sys.time(), '%F')`"
output: hrbrthemes::ipsum_pdf
params:
  school: "Canyon Crest Academy"
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,  dev = "cairo_pdf")
```

```{r libraries, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse) # 
library(tidyr) #
library(here) #
library(fivethirtyeight) #
library(extrafont) #
library(ggrepel) #
library(ggpubr) #
library(hrbrthemes)
loadfonts(device = "win")
```

```{r load-data, include=FALSE}
nsc20.1 <- read.csv(file = here("src/nsc20.1.csv"))
```

# Overview

Every year hundreds of San Diego high schools graduate more than 30,000 seniors.
Many of these graduates go on to enroll in college and attain a degree. The San
Diego County Office of Education's (SDCOE) State of College Report provides
information for schools who share data with the National Student Clearinghouse
(NSC) to answer the following questions:

1.  Where are students going to college?
2.  What percentage of students enroll in college?
3.  What percent of students are retained by these colleges?
4.  What percent complete a degree in six years?

## Data Background

SDCOE downloads data from the NSC annually on February 15. To receive a school
report, high schools must upload an error-free graduates file. SDCOE pulls the
data for these reports directly from the NSC. Any errors in a district's
graduates file contributes to errors contained in these reports. School level
errors also affect county averages.

# State of College Report

In 2020, local high schools reported approximately 26,514 high school graduates
to the NSC; the NSC matched 15,964 of these graduates to enrollment records at
over 1900 colleges and universities. NSC records fluctuate annually depending on
the districts who upload data to the NSC by November. The following graphs
depict where these students enrolled, the percent enrolled, the percent of
students from the class of 2019 who stayed at the college where they first
enrolled, and the percent of students from the classes of 2013 and 2014 who
completed a degree.

\newpage

## College Enrollment

Students across San Diego County attend over 1900 higher education institutions.
Most students graduating from `r params$school` attended the following
institutions in 2020.

```{r fig-1-data, include=FALSE}
inst_enroll <- nsc20.1 %>%
  filter(chrt_grad == 2020) %>%
  group_by(school, college_name_y1) %>%
  summarise(num_enrolled = length(college_name_y1)) %>%
  mutate(college_name_y1 = as.factor(college_name_y1),
         school = as.factor(school))

inst_enroll_2 <- inst_enroll %>% 
  drop_na() %>%
  arrange(school,desc(num_enrolled)) %>%
  slice_head(n = 5) %>%
  filter(school == params$school)
```

```{r fig-1,echo=FALSE, fig.width=7, fig.height=3.5, warning=FALSE, message=FALSE}
inst_enroll_2 %>% # wrap this in a ifelse
  {if(nrow(inst_enroll_2) == 0) {
    cat("No data reported in 2020")
  } else{
  ggdotchart(inst_enroll_2, x = "college_name_y1", y = "num_enrolled",
             group = "college_name_y1", color = "college_name_y1",
             sorting = "descending",
             palette = c("#30a2da","#fc4f30", "#e5ae38", "#6d904f", "#8b8b8b"),
             rotate = TRUE,                                # Rotate vertically
             dot.size = 4,                                 # Large dot size
             y.text.col = FALSE,
  ) +
  geom_hline(yintercept = 0, size = 0.75, color = "#737373") +
  ggtitle(inst_enroll_2$school) +
  labs(subtitle = "Most frequently attended 2 and 4 year colleges",
       caption = "Source: National Student Clearinghouse") +
  scale_y_continuous(
    name = "# Enrolled",
    limits = c(0, 150),
    breaks = seq(0, 150, by = 20)) +
  ggthemes::theme_fivethirtyeight(base_size = 12, base_family = "Chivo") +
  theme(
    plot.title = element_text(color = "black"),
    plot.subtitle = element_text(),
    plot.caption = element_text(color = "gray50"),
    text = element_text(family = "sans"),
    axis.text.x = element_text(color = "gray50"),
    axis.text.y = element_text(color = "gray50"),
    legend.position = "none"
  )
  }
  }
```

The next graph illustrates the percent of graduates from `r params$school`
enrolled in 2 and 4 year colleges between 2012 to 2020 compared to the average
enrollment rates across San Diego county. San Diego County had a steep decline
in college enrollment during the pandemic.

```{r fig-2-data, include=FALSE}
nsc_enroll <- nsc20.1 %>%
  group_by(school, chrt_grad) %>%
  summarise(enrolled = mean(college_enrolled)) %>%
  mutate(school = as.factor(school),
         chrt_grad = as.factor(chrt_grad)) %>%
  filter(school == params$school)

county_average <- nsc20.1 %>%
  group_by(county, chrt_grad) %>%
  summarise(enrolled = mean(college_enrolled)) %>%
  mutate(county = as.factor(county),
         chrt_grad = as.factor(chrt_grad))
```

```{r fig-2,echo=FALSE, fig.width=7, fig.height=3.5, warning=FALSE, message=FALSE}
nsc_enroll %>%
  ggplot(aes(x = chrt_grad, y = enrolled)) +
  geom_line(
    group = 1, 
    size = 1.5,
    color = "#e5ae38") +
  geom_point(size = 2.5,
             color = "#e5ae38") +
  geom_text_repel(aes(
    label = ifelse(chrt_grad == "2020", scales::percent(enrolled, accuracy = 2L), "")),
    force        = 0.5,
    nudge_x      = 0,
    direction    = "y",
    hjust        = 0,
    segment.size = 0.2,
    color        = "#e5ae38",
    fontface     = "bold",
    show.legend = FALSE) +
  geom_line(county_average,
            mapping = aes(x = chrt_grad, y = enrolled), 
            group = 1,
            size = 1.5,
            color = "#30a2da") +
  geom_point(county_average,
             mapping = aes(x = chrt_grad, y = enrolled),
             size = 2.5,
             color = "#30a2da") +
  geom_text_repel(county_average, mapping = aes(x = chrt_grad, y = enrolled,
                                                label = ifelse(chrt_grad == "2020", 
                                                               scales::percent(
                                                                 enrolled, accuracy = 2L), "")),
                  force        = 0.5,
                  nudge_x      = 0,
                  direction    = "y",
                  hjust        = 0,
                  segment.size = 0.2, 
                  color        = "#30a2da",
                  fontface     = "bold",
                  show.legend = FALSE) +
  geom_hline(yintercept = 0, size = 0.75, color = "#737373") +
  ggtitle(nsc_enroll$school) +
  labs(subtitle = "Percentage of graduates enrolled in 2 and 4 year colleges from 2012 to 2020 \n compared to average enrollment across San Diego county",
       caption = "Source: National Student Clearinghouse") +
  scale_y_continuous(
    name = "Year",
    breaks = seq(.0, 1, by = .2),
    labels = scales::percent_format(accuracy = 1)) +
  ggthemes::theme_fivethirtyeight(base_size = 12, base_family = "Chivo") +
  theme(
    plot.title = element_text(color = "black"),
    plot.caption = element_text(color = "gray50"),
    text = element_text(family = "sans"),
    axis.text.x = element_text(color = "gray50"),
    axis.text.y = element_text(color = "gray50")
  ) +
  annotate(
    geom = "text", x = 9, y = 1.05, label = "County",
    fontface = "bold", vjust = 1, color = "#30a2da") +
  annotate(
    geom = "text", x = 8.1, y = 1.05, label = "School",
    fontface = "bold", vjust = 1, color = "#e5ae38")
```

\newpage

## College Retention

This figure illustrates the percent of students who are retained by their
starting institution. Retention is a key factor in college completion. The more
students who remain at their starting institution, the more students graduate.

```{r fig-3-data, include=FALSE}
reten_schl <- nsc20.1 %>% #provide a 0 if no data
  group_by(school, chrt_grad) %>%
  filter(college_enrolled == TRUE, 
         chrt_grad == 2019) %>%
  mutate(retention = coalesce(college_name_y1, '') == coalesce(college_name_y2, '')) %>%
  summarise(n_retention = sum(retention == TRUE, na.rm = TRUE),
            n_enrolled = sum(college_enrolled == TRUE),
            reten = mean(retention, na.rm = TRUE)) %>%
  mutate(school = as.factor(school),
         chrt_grad = as.factor(chrt_grad)) %>%
    filter(school == params$school)

county_average <- nsc20.1 %>%
  group_by(county, chrt_grad) %>%
  filter(college_enrolled == TRUE, 
         chrt_grad == 2019) %>%
  mutate(retention = coalesce(college_name_y1, '') == coalesce(college_name_y2, ' ')) %>%
  summarise(n_retention = sum(retention == TRUE, na.rm = TRUE),
            n_enrolled = sum(college_enrolled == TRUE),
            reten = mean(retention, na.rm = TRUE)) %>%
  mutate(county = as.factor(county),
         chrt_grad = as.factor(chrt_grad))
```

```{r fig-3, echo=FALSE, fig.height=3.5, fig.width=7, warning=FALSE, message=FALSE}
reten_schl %>% 
  {if(nrow(reten_schl) == 0) {
    cat("No retention data reported in 2019")
  } else{
    ggplot(reten_schl, aes(x = chrt_grad, y = reten)) +
       geom_pointrange(aes(ymin = 0, ymax = reten),
                       color = "#e5ae38", fatten = 5, size = 1.5) +
       geom_pointrange(data = county_average, aes(ymin = 0, ymax = reten),
                       color = "#30a2da", fatten = 5, size = 1.5, position = position_nudge(x = 0.5)) +
       geom_hline(yintercept = 0, size = 0.75, color = "#737373") +
       ggtitle(reten_schl$school) +
       labs(subtitle = "Percentage of students enrolled in 2 and 4 year colleges in 2019 retained by the \n starting instituion compared to the average retention in San Diego county",
            caption = "Source: National Student Clearinghouse") +
       coord_flip() +
       scale_y_continuous(
         name = "Year",
         breaks = seq(.0, 1, by = .2),
         labels = scales::percent_format(accuracy = 1)) +
       ggthemes::theme_fivethirtyeight(base_size = 12, base_family = "Chivo") +
       theme(
         plot.title = element_text(color = "black"),
         plot.caption = element_text(color = "gray50"),
         text = element_text(family = "sans"),
         axis.text.x = element_text(color = "gray50"),
         axis.text.y = element_text(color = "gray50")
       ) +
       annotate(
         geom = "text", x = 0.5, y = 1, label = "County",
         fontface = "bold", vjust = 0, color = "#30a2da") +
       annotate(
         geom = "text", x = 0.5, y = .9, label = "School",
         fontface = "bold", vjust = 0, color = "#e5ae38")
  } 
  }
```

## College Completion

The college graduation rates for the students who completed high school in 2013
and 2014 is about 20% countywide (i.e., a six-year graduation rate). Note:
Colleges report graduation rates many different ways. This percentage might be
higher if we count college graduates who completed high school after 2013 and
2014.

```{r fig-4-data, include=FALSE}
completion <- read.csv(file = here("src/completion.csv"))

college_completion <- nsc20.1 %>%
  #mutate(sid = as.factor(sid)) %>%
  left_join(completion, by = "sid") %>%
  dplyr::select(sid, college_enrolled, chrt_grad, county, district, school,
                enroll_begin_y1, seamless_enroll, graduated, college_grad_date,
                degree_title, major, college_sequence) %>%
  filter(chrt_grad == 2013 | 
    chrt_grad == 2014) %>%
  mutate(graduated = ifelse(graduated == "Y", 1, 0)) %>%
  group_by(county, school, chrt_grad) %>%
  summarise(graduated = mean(graduated, na.rm = TRUE)) %>%
  mutate(school = as.factor(school),
         chrt_grad = as.factor(chrt_grad)) 

college_completion_2 <- college_completion %>%
  filter(school == params$school)

county_college_comp_average <- college_completion %>%
  group_by(county, chrt_grad) %>%
  summarize(graduated = mean(graduated, na.rm = TRUE))
```

```{r fig-4, echo=FALSE, fig.width=7, fig.height=3.5, warning=FALSE, message=FALSE}
college_completion_2 %>%
  ggplot(mapping = aes(x = chrt_grad, y = graduated)) +
  geom_line(
    group = 1, 
    size = 1.5,
    color = "#e5ae38") +
  geom_point(size = 2.5,
             color = "#e5ae38") +
  geom_text_repel(aes(
    label = ifelse(chrt_grad == "2014", scales::percent(graduated, accuracy = 2L), "")),
    force        = 0,
    nudge_x      = 0.1,
    direction    = "y",
    hjust        = 0,
    segment.size = 0.2,
    color        = "#e5ae38",
    fontface     = "bold",
    show.legend = FALSE) +
  geom_line(county_college_comp_average,
            mapping = aes(x = chrt_grad, y = graduated), 
            group = 1,
            size = 1.5,
            color = "#30a2da") +
  geom_point(county_college_comp_average,
             mapping = aes(x = chrt_grad, y = graduated),
             size = 2.5,
             color = "#30a2da") +
  geom_text_repel(county_college_comp_average, mapping = aes(x = chrt_grad, y = graduated,
                                                label = ifelse(chrt_grad == "2014", 
                                                               scales::percent(graduated, accuracy = 2L), "")),
                  force        = 0.5,
                  nudge_x      = 0.2,
                  direction    = "y",
                  hjust        = 0.1,
                  segment.size = 0.2, 
                  color        = "#30a2da",
                  fontface     = "bold",
                  show.legend = FALSE) +
  geom_hline(yintercept = 0, size = 0.75, color = "#737373") +
  ggtitle(college_completion_2$school) +
  labs(subtitle = "Percentage of graduates from classes of 2013 to 2014 compared to average \n graduation rates for these classes across San Diego county",
       caption = "Source: National Student Clearinghouse") +
  scale_y_continuous(
    name = "Year",
    limits = c(0,.6),
    breaks = seq(.0, .6, by = .2),
    labels = scales::percent_format(accuracy = 1)) +
  ggthemes::theme_fivethirtyeight(base_size = 12, base_family = "Chivo") +
  theme(
    plot.title = element_text(color = "black"),
    plot.caption = element_text(color = "gray50"),
    text = element_text(family = "sans"),
    axis.text.x = element_text(color = "gray50"),
    axis.text.y = element_text(color = "gray50")
  ) +
  annotate(
    geom = "text", x = 2.3, y = .55, label = "County",
    fontface = "bold", vjust = 0, color = "#30a2da") +
  annotate(
    geom = "text", x = 2.1, y = .55, label = "School",
    fontface = "bold", vjust = 0, color = "#e5ae38")
```
